---
title: "dataprep: fast reshaping with melt() and dcast()"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: fast reshaping with melt() and dcast()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center"
)
```

```{r setup}
library(dataprep)
set.seed(1)
```

## Why another reshape implementation

`melt()` and `dcast()` in `dataprep` 0.1.6 are drop-in replacements
for `reshape2::melt` / `reshape2::dcast`, but the underlying code
is written in C++ with SIMD (AVX2 / AVX-512) and optional OpenMP
parallelism. On a 1,000,000 x 10 table, `melt()` is about

* 7x faster than `reshape2::melt`,
* 24x faster than `tidyr::pivot_longer`,
* 30x faster than `pandas.melt`.

Benchmarks against the other 7 engines are shipped under
`inst/melt_benchmark.R` and `inst/dcast_benchmark.R`. To reproduce
them:

```{r eval=FALSE}
Sys.setenv(DATAPREP_RUN_BENCHMARK = "1")
source(system.file("melt_benchmark.R", package = "dataprep"))
```

## Wide to long with `melt()`

### Basic usage

```{r}
df <- data.frame(
  id       = 1:3,
  category = factor(c("a", "b", "c")),
  v1       = c(1.1, 2.2, 3.3),
  v2       = c(4.4, 5.5, 6.6)
)
melt(df, id.vars = c("id", "category"))
```

### Measure-side specification

```{r}
melt(df, measure.vars = c("v1", "v2"))
```

### Automatic ID inference

Non-numeric and factor columns are treated as IDs by default.

```{r}
melt(df)
```

### Custom column names and `na.rm`

```{r}
df_na <- data.frame(
  id = 1:3,
  x  = c(1, NA, 3),
  y  = c(4, 5, NA)
)
melt(df_na, id.vars = "id",
     variable.name = "var", value.name = "val",
     na.rm = TRUE)
```

### Memory layout: `major = "row"` vs `major = "col"`

Row-major (`"row"`, default) is usually faster when there are few
measure columns; column-major (`"col"`) is often faster when there
are many measure columns because bulk copies dominate.

```{r}
wide50 <- data.frame(id = 1:100,
                     matrix(rnorm(100 * 50), ncol = 50))
res_row <- melt(wide50, id.vars = "id", major = "row")
res_col <- melt(wide50, id.vars = "id", major = "col")
identical(as.data.frame(res_row), as.data.frame(res_col))
```

### Thread control

`cores` controls OpenMP. `options(dataprep.cores = ...)` sets a
default for the whole session.

```{r eval=FALSE}
options(dataprep.cores = 4L)
melt(df, id.vars = "id")
options(dataprep.cores = NULL)
```

## Long to wide with `dcast()`

### Basic usage

```{r}
long <- melt(df, id.vars = c("id", "category"))
dcast(long, id = c("id", "category"),
      variable = "variable", value = "value")
```

### Formula interface

```{r}
dcast(long, formula = id + category ~ variable,
      value.var = "value")
```

### Fill missing cells

```{r}
dcast(long, id = c("id", "category"),
      variable = "variable", value = "value",
      fill = 0)
```

### Aggregating duplicate pairs

If a `(id, variable)` pair appears more than once, pass
`fun.aggregate`. The default is "last occurrence wins", matching
`data.table::dcast`'s `fun.aggregate = NULL` behaviour.

```{r}
long_dup <- data.frame(
  id       = c(1, 1, 2),
  variable = c("x", "x", "x"),
  value    = c(1, 2, 3)
)
dcast(long_dup, id = "id",
      variable = "variable", value = "value",
      fun.aggregate = mean)
```

### `na.rm`

```{r}
long_na <- data.frame(
  id       = c(1, 1, 2, 2),
  variable = c("x", "y", "x", "y"),
  value    = c(1, NA, 3, 4)
)
dcast(long_na, id = "id",
      variable = "variable", value = "value",
      na.rm = TRUE)
```

## Consistency with other engines

`dataprep::melt` and `dcast` produce the same output as
`reshape2`, `data.table`, `tidyr`, `pandas`, `polars`, `dask`, and
`duckdb` on a wide range of shapes and sizes. The consistency
scripts are shipped under `inst/melt_consistency.R` and
`inst/dcast_consistency.R`. To run them:

```{r eval=FALSE}
Sys.setenv(DATAPREP_RUN_BENCHMARK = "1")
source(system.file("melt_consistency.R",  package = "dataprep"))
source(system.file("dcast_consistency.R", package = "dataprep"))
melt_all_engines(10000L, n_id = 1L, n_val = 9L)
dcast_all_engines(1000L, n_id = 2L, n_val = 5L)
```

## Round-trip example

```{r}
wide  <- data.frame(id = 1:5, a = rnorm(5), b = rnorm(5))
long  <- melt(wide, id.vars = "id")
back  <- dcast(long, id = "id",
               variable = "variable", value = "value")
all.equal(as.data.frame(back)[order(back$id), c("a", "b")],
          wide[, c("a", "b")],
          tolerance = 1e-12)
```

## Session info

```{r}
sessionInfo()
```