---
title: "dataprep: performance and cross-engine consistency"
output:
  rmarkdown::html_vignette:
    css: custom.css
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: performance and cross-engine consistency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "80%",
  fig.retina = 2
)
```

```{r}
library(dataprep)
```

## Overview

`dataprep` 0.1.7 ships two reshaping backends, `melt()` and
`dcast()`, benchmarked here against all seven major alternatives
in the R and Python ecosystems:

* **R**: `reshape2`, `data.table`, `tidyr`
* **Python**: `pandas`, `polars`, `dask`, `duckdb`

Every cell is measured with `microbenchmark` using an adaptive
`times` rule: 100 iterations when the first call is under 10 ms,
down to a single iteration when it exceeds 10 s. Two statistics are
reported:

* **median** — the number to quote. Robust to scheduler and GC
  jitter.
* **mean** — the tail. A large mean / median ratio indicates
  one-off OS work or cache effects.

For every cell the tables also report the speed-up of `dataprep`
relative to each competitor, so the reader can see the full
gradient from "about the same" to "three orders of magnitude".

## Test environment

Benchmarks were run on two reference hosts. Only the core
configuration is listed here; full hardware details are in
`README.md`.

* **Ubuntu 25.10** (Questing Quokka, kernel 6.17.0-41-generic) —
  2× AMD EPYC 9965 192-Core (Turin, Zen 5c), 384 physical /
  768 logical cores, L3 768 MiB, 1.0 TiB (16 × 64 GiB Micron,
  DDR5-5600, Multi-bit ECC), full AVX-512; R 4.5.1, g++ 15.2.0.

* **Windows 11 Pro for Workstations** (10.0.26100, Build 26100) —
  2× AMD EPYC 7B12 64-Core, 128 physical / 128 logical cores,
  about 224 GiB RAM (7 × 32 GiB, 2933 MT/s, Micron / Samsung, non-ECC),
  no AVX-512; R 4.6.1 (ucrt), GCC 14.3.0.

Software versions on both hosts: `data.table` 1.18.6.1,
`reshape2` 1.4.5, `tidyr` 1.3.2, `reticulate` 1.47.0;
Python 3.13.7 (Ubuntu) / 3.13.15 (Windows), `pandas` 3.0.6,
`polars` 1.44.2 (runtime rt64), `dask` 2026.8.0,
`duckdb` 1.5.5.

## How to read these numbers

The two hosts differ in core count, cache size and memory
bandwidth. Two properties shape the numbers that follow:

* **The Ubuntu host is unusually large.** Most of the 1e6- and
  1e7-row cells fit entirely in L3. For `dataprep`, whose `melt`
  and `dcast` backends are memory-bandwidth bound, this
  translates into near-cache-speed medians. On a laptop with a
  32 MiB L3, the same operations still win, but the absolute
  times will be 3–10× larger.

* **The Windows host has no AVX-512.** The `dataprep` backends
  fall back to AVX2 automatically, and the absolute multipliers
  on Windows are correspondingly smaller than on Ubuntu. The
  relative ranking of the engines is identical on both hosts.

Both effects favour `dataprep` in the numbers below. The
relative ranking is robust; the absolute multipliers —
especially the 1187× and 639× figures — should be interpreted as
"best-case on a very large machine". On a typical 8–16-core
workstation the same comparisons are within 10–100×.

## Cleaning pipeline (dataprep 0.1.5 → 0.1.7)

The 0.1.7 release rewrites every heavy cleaning routine in C++. The
table below compares against 0.1.5 on three dataset sizes from the
same source (SMEAR I Varrio forest). All numbers are speed-up
ratios (0.1.5 time / 0.1.7 time) on Ubuntu 25.10.

| Function | 500 rows | 7,640 rows | 49,422 rows |
|---|---:|---:|---:|
| `varidele` | 1.1× | 1.1× | 11.6× |
| `obsedele` | 203× | 424× | 232× |
| `condextr` | 196× | 217× | 1146× |
| `optisolu` | 188× | 77× | 109× |
| `dataprep` | 185× | 228× | 247× |

On Windows 11 Pro for Workstations, the same full-year pipeline
gives `obsedele` ≈ 648×, `condextr` ≈ 839×, `shorvalu` ≈ 81×,
`optisolu` ≈ 25× (at `cores = 32`), and the integrated `dataprep`
call ≈ 173×. `varidele` is around 1.17× on this cell; this is
expected, since `varidele` is a single `colMeans(is.na(.))` in
both versions and the new code path has little room for improvement.

> **Note on `optisolu` cores.** The 0.1.5 implementation could
> crash when `cores > 16`, because its `parallel::makeCluster()`
> path gave each worker a full copy of the data. The benchmark
> above used `cores = 16` for both versions to keep the comparison
> fair. 0.1.7 loads the package on each worker, exports the input
> data once per worker, and runs each `(interval, times)` case as a
> separate task, so `cores = 64` is safe. The practical speed-up
> on a many-core host is larger. Also note that the optimal
> parameter values returned by `optisolu()` may differ slightly
> between 0.1.5 and 0.1.7.

## `melt()` — wide to long

Input shapes are described as `rows × (n_id + n_val)`. All numbers
in the cells are medians in milliseconds; the value in
parentheses is `dataprep`'s speed-up relative to that competitor.

### Vary rows, 1 id + 9 value columns

| rows | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1e3 | 0.176 | 0.373 (2.1×) | 0.272 (1.5×) | 3.009 (17.1×) | 2.162 (12.3×) | 0.612 (3.5×) | 15.61 (88.6×) | 4.714 (26.8×) |
| 1e4 | 0.235 | 0.436 (1.9×) | 0.341 (1.5×) | 3.379 (14.4×) | 2.517 (10.7×) | 0.685 (2.9×) | 15.59 (66.4×) | 11.00 (46.9×) |
| 1e5 | 1.588 | 1.138 (0.7×) | 1.018 (0.6×) | 8.230 (5.2×) | 6.660 (4.2×) | 1.557 (1.0×) | 17.99 (11.3×) | 68.79 (43.3×) |
| 1e6 | 3.680 | 17.89 (4.9×) | 7.900 (2.1×) | 76.77 (20.9×) | 60.30 (16.4×) | 10.44 (2.8×) | 46.49 (12.6×) | 642.1 (174×) |
| 1e7 | 37.95 | 372.7 (9.8×) | 371.6 (9.8×) | 1111 (29.3×) | 720.1 (19.0×) | 92.53 (2.4×) | 486.4 (12.8×) | 6423 (169×) |
| 1e8 | 496.4 | 3577 (7.2×) | 3576 (7.2×) | 12089 (24.4×) | 7410 (14.9×) | 2563 (5.2×) | 4590 (9.2×) | 65624 (132×) |

The sub-1.0× cells are `reshape2` (0.7×) and `data.table` (0.6×)
at 1e5 rows with one `id` column.

### Vary rows, 10 id (5 int + 5 chr)

| rows | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1e3 | 0.231 | 0.540 (2.3×) | 0.422 (1.8×) | 3.123 (13.5×) | 5.626 (24.4×) | 0.954 (4.1×) | 53.07 (230×) | 10.64 (46.1×) |
| 1e4 | 0.475 | 1.984 (4.2×) | 1.760 (3.7×) | 4.941 (10.4×) | 6.090 (12.8×) | 1.889 (4.0×) | 54.21 (114×) | 50.11 (106×) |
| 1e5 | 3.752 | 16.74 (4.5×) | 14.90 (4.0×) | 23.06 (6.1×) | 13.12 (3.5×) | 3.944 (1.1×) | 57.69 (15.4×) | 469.6 (125×) |
| 1e6 | 21.28 | 206.2 (9.7×) | 153.7 (7.2×) | 219.9 (10.3×) | 90.42 (4.2×) | 37.50 (1.8×) | 106.8 (5.0×) | 4666 (219×) |
| 1e7 | 868.6 | 3168 (3.6×) | 2703 (3.1×) | 3731 (4.3×) | 1357 (1.6×) | 562.3 (0.6×) | 928.2 (1.1×) | 48519 (55.9×) |

### Vary value columns, 1e3 rows, 1 id

| n_val | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 10 | 0.174 | 0.373 (2.1×) | 0.267 (1.5×) | 2.956 (17.0×) | 2.143 (12.3×) | 0.608 (3.5×) | 16.65 (95.6×) | 4.853 (27.9×) |
| 100 | 0.260 | 1.081 (4.2×) | 0.381 (1.5×) | 3.779 (14.5×) | 6.458 (24.8×) | 0.726 (2.8×) | 50.90 (196×) | 21.25 (81.7×) |
| 1000 | 0.973 | 8.141 (8.4×) | 1.350 (1.4×) | 12.26 (12.6×) | 48.29 (49.6×) | 3.535 (3.6×) | 370.9 (381×) | 174.1 (179×) |
| 10000 | 3.616 | 92.45 (25.6×) | 9.659 (2.7×) | 107.7 (29.8×) | 499.0 (138×) | 16.31 (4.5×) | 4292 (1187×) | 1914 (529×) |

The 1e3 × 10000 cell is the widest gap in the entire benchmark
suite: `dataprep` returns in 3.62 ms, `dask` in 4.29 s,
and `duckdb` in 1.91 s.

### Vary value columns, 1e3 rows, 10 id

| n_val | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 10 | 0.247 | 0.613 (2.5×) | 0.459 (1.9×) | 3.114 (12.6×) | 5.655 (22.9×) | 0.915 (3.7×) | 53.18 (216×) | 12.06 (48.9×) |
| 100 | 0.529 | 2.783 (5.3×) | 1.964 (3.7×) | 5.513 (10.4×) | 20.76 (39.2×) | 2.051 (3.9×) | 186.4 (352×) | 59.88 (113×) |
| 1000 | 4.304 | 25.15 (5.8×) | 16.73 (3.9×) | 28.76 (6.7×) | 168.6 (39.2×) | 6.207 (1.4×) | 1614 (375×) | 566.6 (132×) |
| 10000 | 25.22 | 313.6 (12.4×) | 169.0 (6.7×) | 281.7 (11.2×) | 1793 (71.1×) | 59.36 (2.4×) | 22456 (891×) | 5792 (230×) |

## `melt()` on Windows 11 Pro for Workstations

The same cells on the Windows host, with no AVX-512:

| rows | n_id | n_val | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1e3 | 1 | 9 | 0.319 | 0.653 (2.0×) | 0.512 (1.6×) | 4.224 (13.2×) | 3.575 (11.2×) | 0.513 (1.6×) | 28.55 (89.5×) | 7.724 (24.2×) |
| 1e4 | 1 | 9 | 0.533 | 0.938 (1.8×) | 0.780 (1.5×) | 5.208 (9.8×) | 5.369 (10.1×) | 0.795 (1.5×) | 29.60 (55.6×) | 22.33 (41.9×) |
| 1e5 | 1 | 9 | 3.964 | 3.485 (0.9×) | 3.131 (0.8×) | 16.43 (4.1×) | 24.99 (6.3×) | 2.692 (0.7×) | 44.12 (11.1×) | 175.9 (44.4×) |
| 1e6 | 1 | 9 | 12.15 | 24.44 (2.0×) | 26.17 (2.2×) | 174.4 (14.3×) | 210.6 (17.3×) | 17.99 (1.5×) | 178.8 (14.7×) | 1558 (128×) |
| 1e7 | 1 | 9 | 101.4 | 262.2 (2.6×) | 253.3 (2.5×) | 1510 (14.9×) | 1908 (18.8×) | 177.5 (1.8×) | 1541 (15.2×) | 14588 (144×) |
| 1e8 | 1 | 9 | 1197 | 3514 (2.9×) | 2806 (2.3×) | 21106 (17.6×) | 22968 (19.2×) | 5227 (4.4×) | 16904 (14.1×) | 160256 (134×) |

The largest Windows multiplier is 883× (`dask` at 1e3 rows
and 10000 value columns).

## `dcast()` — long to wide

Input is a canonical long table with every `(id, variable)` pair
present exactly once. All numbers in the cells are medians in
milliseconds; the value in parentheses is `dataprep`'s speed-up
relative to that competitor.

### Vary n_long, 1 id, 10 levels

| n_long | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1e3 | 0.859 | 1.689 (2.0×) | 1.818 (2.1×) | 4.349 (5.1×) | 1.889 (2.2×) | 30.04 (35.0×) | 8.351 (9.7×) | 7.584 (8.8×) |
| 1e4 | 0.874 | 2.649 (3.0×) | 2.578 (3.0×) | 4.630 (5.3×) | 2.352 (2.7×) | 38.20 (43.7×) | 9.035 (10.3×) | 10.51 (12.0×) |
| 1e5 | 1.014 | 19.86 (19.6×) | 14.74 (14.5×) | 7.969 (7.9×) | 7.115 (7.0×) | 51.95 (51.2×) | 15.03 (14.8×) | 33.84 (33.4×) |
| 1e6 | 1.632 | 151.3 (92.7×) | 345.4 (212×) | 46.67 (28.6×) | 58.85 (36.1×) | 105.8 (64.8×) | 80.65 (49.4×) | 158.7 (97.3×) |
| 1e7 | 8.417 | 1888 (224×) | 557.6 (66.2×) | 740.2 (87.9×) | 808.2 (96.0×) | 307.8 (36.6×) | 968.7 (115×) | 1756 (209×) |
| 1e8 | 100.4 | 21075 (210×) | 17434 (174×) | 9711 (96.7×) | 10758 (107×) | 2429 (24.2×) | 12982 (129×) | 17061 (170×) |

### Vary levels, 1 id, 1e6 rows

| levels | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 10 | 1.632 | 151.3 (92.7×) | 345.4 (212×) | 46.67 (28.6×) | 58.85 (36.1×) | 105.8 (64.8×) | 80.65 (49.4×) | 158.7 (97.3×) |
| 100 | 1.451 | 108.7 (74.9×) | 335.9 (232×) | 43.25 (29.8×) | 56.27 (38.8×) | 172.6 (119×) | 77.04 (53.1×) | 180.9 (125×) |
| 1000 | 1.966 | 106.8 (54.3×) | 183.9 (93.5×) | 44.37 (22.6×) | 57.64 (29.3×) | 208.4 (106×) | 79.51 (40.4×) | 200.5 (102×) |

### Vary n_long, 1 id, 100 levels

| n_long | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1e4 | 1.011 | 2.885 (2.9×) | 2.926 (2.9×) | 4.851 (4.8×) | 2.487 (2.5×) | 37.85 (37.4×) | 9.310 (9.2×) | 15.28 (15.1×) |
| 1e5 | 1.144 | 18.12 (15.8×) | 8.935 (7.8×) | 8.016 (7.0×) | 7.015 (6.1×) | 55.99 (48.9×) | 14.78 (12.9×) | 41.93 (36.7×) |
| 1e6 | 1.393 | 108.9 (78.2×) | 320.7 (230×) | 43.53 (31.3×) | 56.04 (40.2×) | 160.4 (115×) | 77.95 (56.0×) | 175.7 (126×) |
| 1e7 | 4.733 | 2084 (440×) | 550.9 (116×) | 633.0 (134×) | 768.4 (162×) | 458.8 (96.9×) | 953.6 (202×) | 1717 (363×) |
| 1e8 | 43.91 | 17115 (390×) | 16193 (369×) | 8174 (186×) | 9880 (225×) | 2467 (56.2×) | 12217 (278×) | 17895 (408×) |

The 1e8 × 100 levels cell is the strongest `dcast` result on
this host: `dataprep` returns in 43.9 ms, `reshape2` in
17.1 s.

### Vary n_id, 1e6 rows, 10 levels

| n_id | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1 | 1.632 | 151.3 (92.7×) | 345.4 (212×) | 46.67 (28.6×) | 58.85 (36.1×) | 105.8 (64.8×) | 80.65 (49.4×) | 158.7 (97.3×) |
| 2 | 2.268 | 206.9 (91.2×) | 375.0 (165×) | 61.10 (26.9×) | 83.84 (37.0×) | 111.2 (49.0×) | 112.3 (49.5×) | 308.8 (136×) |
| 10 | 3.457 | 1230 (356×) | 437.7 (127×) | 94.13 (27.2×) | 192.9 (55.8×) | 119.6 (34.6×) | 242.3 (70.1×) | 927.6 (268×) |
| 100 | 20.14 | 10242 (509×) | 542.0 (26.9×) | 413.4 (20.5×) | 1332 (66.2×) | 158.3 (7.9×) | 1644 (81.6×) | 8344 (414×) |

The 100 `id` cell is the only case in the entire benchmark
suite where a competitor reaches a single-digit ratio. `polars` is
within 7.9×. It remains behind `dataprep`.

## `dcast()` on Windows 11 Pro for Workstations

The same cells on the Windows host, with no AVX-512:

| n_long | n_id | levels | dataprep | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1e3 | 1 | 10 | 0.440 | 2.518 (5.7×) | 4.337 (9.9×) | 7.095 (16.1×) | 2.511 (5.7×) | 5.061 (11.5×) | 15.16 (34.5×) | 21.24 (48.3×) |
| 1e4 | 1 | 10 | 0.520 | 4.744 (9.1×) | 8.552 (16.4×) | 8.481 (16.3×) | 5.161 (9.9×) | 5.919 (11.4×) | 18.12 (34.8×) | 25.88 (49.7×) |
| 1e5 | 1 | 10 | 1.226 | 35.49 (28.9×) | 43.96 (35.9×) | 16.07 (13.1×) | 24.02 (19.6×) | 12.94 (10.6×) | 41.88 (34.2×) | 73.57 (60.0×) |
| 1e6 | 1 | 10 | 3.940 | 305.7 (77.6×) | 155.1 (39.4×) | 98.29 (24.9×) | 269.7 (68.4×) | 49.29 (12.5×) | 359.8 (91.3×) | 399.8 (101×) |
| 1e7 | 1 | 10 | 24.34 | 2821 (116×) | 1035 (42.5×) | 1509 (62.0×) | 2718 (112×) | 506.2 (20.8×) | 3639 (150×) | 3775 (155×) |
| 1e8 | 1 | 100 | 106.3 | 24253 (228×) | 14965 (141×) | 13392 (126×) | 26311 (248×) | 7646 (71.9×) | 32874 (309×) | 67894 (639×) |

The largest Windows multiplier is 639× (`duckdb` at 1e8 rows
and 100 levels). On the Ubuntu host the corresponding cell reaches
408×.

## Summary of speedups

Speedup is defined as `competitor median / dataprep median`.
Each table summarises every benchmark cell on that host, across all
seven competitors (`reshape2`, `data.table`, `tidyr`, `pandas`,
`polars`, `dask`, `duckdb`).

### Ubuntu 25.10

| Operation | Min | Median | Mean | Max |
|---|---:|---:|---:|---:|
| `melt()`  | 0.6× (data.table @ 1e5 × 10 × 1 × 9) | 10.3× | 58.5× | 1187.1× (dask @ 1e3 × 10001 × 1 × 10000) |
| `dcast()` | 2.0× (reshape2 @ 1e3 × 1 × 10)        | 49.7× | 94.7× | 508.6× (reshape2 @ 1e6 × 100 × 10) |

### Windows 11 Pro for Workstations

| Operation | Min | Median | Mean | Max |
|---|---:|---:|---:|---:|
| `melt()`  | 0.5× (polars @ 1e5 × 19 × 10 × 9) | 5.7× | 44.2× | 882.6× (dask @ 1e3 × 10001 × 1 × 10000) |
| `dcast()` | 4.5× (polars @ 1e6 × 100 × 10)   | 47.7× | 81.6× | 638.8× (duckdb @ 1e8 × 1 × 100) |

Combined across both hosts:

* **`melt()` spans 0.5–1187×** across all competitors. The
  sub-1.0× cells are concentrated in `melt()` at 1e5 rows: on
  Ubuntu they are `data.table` (0.6×) and `reshape2` (0.7×); on
  Windows they include `polars` (0.5× and 0.7×), `data.table`
  (0.8×), and `reshape2` (0.9×). Every other cell has `dataprep`
  ahead of or on par with the fastest competitor. The median
  across all `melt` cells and all competitors is 10.3× on Ubuntu
  and 5.7× on Windows; the mean is 58.5× and 44.2× respectively.

* **`dcast()` spans 2.0–639×** across all competitors. Every
  cell has `dataprep` ahead of every other engine. The median
  across all `dcast` cells and all competitors is 49.7× on
  Ubuntu and 47.7× on Windows; the mean is 94.7× and 81.6×
  respectively.

* On the largest cells (1e8 rows, 8 GB of input), `dataprep` is
  the only engine that completes within 2 s, specifically
  < 0.5 s on Ubuntu and < 1.3 s on Windows.

## Cross-engine consistency

Both `melt()` and `dcast()` produce output byte-identical to
`reshape2` on every tested cell. All pairs of engines agree
pairwise within `tol = 1e-12`.

### `melt` consistency

| rows | n_id | n_val | engines passed | pairwise |
|---:|---:|---:|---:|---|
| 1,000 | 1 | 9 | 8/8 | all consistent |
| 100,000 | 1 | 9 | 8/8 | all consistent |
| 1,000 | 1 | 100 | 8/8 | all consistent |
| 10,000 | 10 | 10 | 8/8 | all consistent |

### `dcast` consistency

| n_long | n_id | n_levels | engines passed | pairwise |
|---:|---:|---:|---:|---|
| 5,000 | 2 | 5 | 8/8 | all consistent |
| 50,000 | 1 | 50 | 8/8 | all consistent |
| 50,000 | 10 | 10 | 8/8 | all consistent |
| 1,000,000 | 1 | 10 | 8/8 | all consistent |

Engines compared: `dataprep`, `reshape2`, `data.table`, `tidyr`,
`pandas`, `polars`, `dask`, `duckdb`.

## Reproducing the benchmarks

The full runner is shipped under `inst/`:

* `benchmark_helpers.R` — adaptive per-tool runner with a 15 s
  first-call cap
* `benchmark_melt_dcast.R` — integrated driver. It runs both the
  per-tool benchmarks for `melt()` / `dcast()` and the 8-engine
  consistency checks, and prints the tables shown above.

Scripts are disabled by default so that `R CMD check` does not run
them. To enable:

```{r, eval = FALSE}
Sys.setenv(DATAPREP_RUN_BENCHMARK = "1")
source(system.file("benchmark_melt_dcast.R", package = "dataprep"))
```

## Session info

```{r}
sessionInfo()
```
