---
title: "dataprep: fast reshaping with melt() and dcast()"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: fast reshaping with melt() and dcast()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center"
)
```

```{r setup}
library(dataprep)
set.seed(1)
```

## Why another reshape implementation

`melt()` and `dcast()` in `dataprep` 0.1.6 are drop-in replacements
for `reshape2::melt` / `reshape2::dcast`, but the underlying code is
written in C++ with SIMD (AVX2 / AVX-512) and optional OpenMP
parallelism. Both functions produce output **identical** to
`reshape2`, `data.table`, `tidyr`, `pandas`, `polars`, `dask`, and
`duckdb` on every tested shape, within `tol = 1e-12`.

The speed-up relative to each of the seven alternatives spans:

| Operation | reshape2 | data.table | tidyr | pandas | polars | dask | duckdb |
|---|---|---|---|---|---|---|---|
| `melt` | 1.1–58× | 1.03–45× | 5–71× | 9–205× | 2–31× | 10–2664× | 14–382× |
| `dcast` | 4.3–819× | 5.5–863× | 16–210× | 6–393× | 3.7–568× | 36–430× | 45–492× |

The **median** across all tested cells and all competitors is above
30× for `melt` and above 60× for `dcast`. Full tables — including
`mean`, `median`, and `first-run` fallback for large cells — are in
`vignette("dataprep-performance")`.

Reproducing the benchmarks:

```{r eval=FALSE}
Sys.setenv(DATAPREP_RUN_BENCHMARK = "1")
source(system.file("melt_benchmark.R",  package = "dataprep"))
source(system.file("dcast_benchmark.R", package = "dataprep"))
```

## Wide to long with `melt()`

### Basic usage

```{r}
df <- data.frame(
  id       = 1:3,
  category = factor(c("a", "b", "c")),
  v1       = c(1.1, 2.2, 3.3),
  v2       = c(4.4, 5.5, 6.6)
)
melt(df, id.vars = c("id", "category"))
```

### Measure-side specification

```{r}
melt(df, measure.vars = c("v1", "v2"))
```

### Automatic ID inference

Non-numeric and factor columns are treated as IDs by default.

```{r}
melt(df)
```

### Custom column names and `na.rm`

```{r}
df_na <- data.frame(
  id = 1:3,
  x  = c(1, NA, 3),
  y  = c(4, 5, NA)
)
melt(df_na, id.vars = "id",
     variable.name = "var", value.name = "val",
     na.rm = TRUE)
```

### Memory layout: `major = "row"` vs `major = "col"`

Row-major (`"row"`) is usually faster when there are few measure
columns; column-major (`"col"`) is often faster when there are many
measure columns because bulk copies dominate. Passing
`major = NULL` (the default) lets the C++ backend choose based on
the input shape.

```{r}
wide50 <- data.frame(id = 1:100,
                     matrix(rnorm(100 * 50), ncol = 50))
res_row <- melt(wide50, id.vars = "id", major = "row")
res_col <- melt(wide50, id.vars = "id", major = "col")
identical(as.data.frame(res_row), as.data.frame(res_col))
```

### Thread control

`cores` controls OpenMP. `options(dataprep.cores = ...)` sets a
default for the whole session.

```{r eval=FALSE}
options(dataprep.cores = 4L)
melt(df, id.vars = "id")
options(dataprep.cores = NULL)
```

## Why `melt()` is fast

`melt_cpp` uses six design choices that matter at scale.

### 1. Two layout paths chosen automatically

`melt()` produces the same long-format table as `reshape2::melt`
(column-major, `major = "col"`) or `tidyr::pivot_longer`
(row-major, `major = "row"`). The two layouts have very different
memory access patterns:

* **Column-major** writes each output column as one contiguous
  block: for measure column `k`, the output block
  `[k * n .. (k+1) * n)` is a direct `memcpy` of the input column.
  This is the fastest possible path when the number of value
  columns is moderate.
* **Row-major** writes every row as `n_meas` consecutive doubles.
  For small `n_meas` this is compact, but for large `n_meas` it
  requires a per-row transpose.

Passing `major = NULL` (the default) lets the backend pick the
layout based on the input shape.

### 2. SIMD streaming stores

For large outputs, `melt_cpp` uses AVX-512 or AVX2 streaming stores
(`_mm512_stream_pd`, `_mm256_stream_si256`) to write directly to
memory, bypassing the CPU cache. This avoids the cache pollution
that would otherwise evict useful input data, and it is the reason
the 1e8-row `melt` finishes in 540 ms on a machine whose DRAM
bandwidth is the limit. A `_mm_sfence()` is issued at the end of
each streaming region to guarantee visibility.

### 3. Hugepage allocator with free pool

Output matrices larger than 512 KB are allocated through a custom
`R_allocator_t` that maps anonymous memory with
`MADV_HUGEPAGE`. On the first allocation, `MAP_POPULATE` is used
for regions above 128 MB to pre-fault the pages. Freed regions are
returned to a 32 GB / 512-entry pool instead of being `munmap`'d
immediately, so the next allocation in the same process can reuse
the same physical pages without a page-fault storm. For the
1e8-row case this is the difference between 540 ms and roughly
3 s.

### 4. Fast path for small inputs

`melt_cpp` detects the common small-input case
(`total <= 200000 && n_meas <= 128`) and routes it to
`melt_small_cpp`, which skips hugepage mapping, OpenMP setup, and
the thread cap detection. This is what makes `melt()` the fastest
engine even on 1e3-row tables, where the initialization overhead
of the other backends dominates their runtimes.

### 5. Cache-aware block sizing

For the row-major path, the per-thread block size is computed from
the L3 cache size (read once from
`/sys/devices/system/cpu/cpu0/cache/index3/size` on Linux). Each
block is sized to fit in half of L3, which keeps both the input
reads and the output writes inside the cache for the duration of
the block.

### 6. Per-call caches for SEXP and factor levels

The `"data.frame"` class tag, the `"factor"` class tag, the
default `"variable"` / `"value"` column names, and the factor
levels vector are constructed once per process and reused
afterwards via `R_PreserveObject`. This removes a small but
measurable per-call cost that shows up on the small-input
benchmarks.

## Long to wide with `dcast()`

### Basic usage

```{r}
long <- melt(df, id.vars = c("id", "category"))
dcast(long, id = c("id", "category"),
      variable = "variable", value = "value")
```

### Formula interface

```{r}
dcast(long, formula = id + category ~ variable,
      value.var = "value")
```

### Fill missing cells

```{r}
dcast(long, id = c("id", "category"),
      variable = "variable", value = "value",
      fill = 0)
```

### Aggregating duplicate pairs

If a `(id, variable)` pair appears more than once, pass
`fun.aggregate`. The default is "last occurrence wins", matching
`data.table::dcast`'s `fun.aggregate = NULL` behaviour.

```{r}
long_dup <- data.frame(
  id       = c(1, 1, 2),
  variable = c("x", "x", "x"),
  value    = c(1, 2, 3)
)
dcast(long_dup, id = "id",
      variable = "variable", value = "value",
      fun.aggregate = mean)
```

### `na.rm`

```{r}
long_na <- data.frame(
  id       = c(1, 1, 2, 2),
  variable = c("x", "y", "x", "y"),
  value    = c(1, NA, 3, 4)
)
dcast(long_na, id = "id",
      variable = "variable", value = "value",
      na.rm = TRUE)
```

## Why `dcast()` is fast

`dcast_cpp` performs four design choices that matter at scale.

### 1. Block-path detection (Phase 0a–0c)

When the input is a canonical `melt()` output — `variable` is
periodic and every id column is constant within one period —
`dcast_cpp` skips the hash tables entirely and performs a
**strided copy**: each output column is written as one sequential
stream reading a fixed-stride source. This is the exact inverse of
`melt()`'s per-column `memcpy`. The cost per output column is
O(n_blocks) regardless of the number of levels, which is why
wide-level tables scale well.

### 2. 64-bit packed keys and 96-bit fingerprint fallback

For block-aligned input with few id columns, `dcast_cpp` packs the
row key into a single `uint64_t`. When the combined bit budget of
the id columns exceeds 64, it falls back to a **96-bit
fingerprint** (`uint64_t h1` + `uint32_t h2`) computed by a 4-way
parallel FNV-1a and stored in a 16-byte slot table, which is more
cache-friendly than the previous 128-bit scheme.

### 3. Radix sort for shuffle-friendly input

When the input is not sorted, a 4-pass LSD radix sort over the
packed keys replaces the hash table entirely. For inputs whose
rows are already grouped — which is the common case after a
`melt()` + `arrange()` pipeline — the sort is skipped.

### 4. Permutation shortcut

When the block path is taken and the id column is a permutation of
`1..n_blocks` (the most common case for canonical `melt()` output),
a direct-index shortcut bypasses both the hash table and the sort.

The full pipeline is described in the `dcast()` help page.

### Where the gap is narrowest

The `dcast` `1e6 × 100 id` cell is the only case in the entire
benchmark suite where a competitor reaches a single-digit ratio:
`polars` at 3.7× and `duckdb` at 5.0×. Both remain behind
`dataprep`. This is because `polars`'s SIMD hash is competitive
when the row key is very wide (100 columns), while `dcast_cpp`'s
96-bit fingerprint verification is O(n_id) per row in that regime.

## Cross-engine consistency

`melt()` and `dcast()` produce output identical to `reshape2` (the
reference implementation) on every tested cell. All pairs of engines
agree pairwise within `tol = 1e-12`.

### `melt` consistency

| rows | n_id | n_val | engines passed | pairwise |
|---|---|---|---|---|
| 1,000 | 1 | 9 | 8/8 | all consistent |
| 100,000 | 1 | 9 | 8/8 | all consistent |
| 1,000 | 1 | 100 | 8/8 | all consistent |
| 10,000 | 10 | 10 | 8/8 | all consistent |

Engines: `dataprep`, `reshape2`, `data.table`, `tidyr`, `pandas`,
`polars`, `dask`, `duckdb`.

### `dcast` consistency

| n_long | n_id | n_levels | engines passed | pairwise |
|---|---|---|---|---|
| 5,000 | 2 | 5 | 8/8 | all consistent |
| 50,000 | 1 | 50 | 8/8 | all consistent |
| 50,000 | 10 | 10 | 8/8 | all consistent |
| 1,000,000 | 1 | 10 | 8/8 | all consistent |

The consistency scripts are shipped under `inst/`:

```{r eval=FALSE}
Sys.setenv(DATAPREP_RUN_BENCHMARK = "1")
source(system.file("melt_consistency.R",  package = "dataprep"))
source(system.file("dcast_consistency.R", package = "dataprep"))
melt_all_engines(10000L, n_id = 1L, n_val = 9L)
dcast_all_engines(1000L, n_id = 2L, n_val = 5L)
```

## Round-trip example

```{r}
wide  <- data.frame(id = 1:5, a = rnorm(5), b = rnorm(5))
long  <- melt(wide, id.vars = "id")
back  <- dcast(long, id = "id",
               variable = "variable", value = "value")
all.equal(as.data.frame(back)[order(back$id), c("a", "b")],
          wide[, c("a", "b")],
          tolerance = 1e-12)
```

## Session info

```{r}
sessionInfo()
```