---
title: "dataprep: cleaning pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: cleaning pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center"
)
```

```{r setup}
library(dataprep)
set.seed(1)
```

## Overview

The 0.1.6 cleaning pipeline consists of four sequential steps:

1. **Variable deletion** (`varidele`): drop columns whose missing
   fraction is above a threshold.
2. **Observation deletion** (`obsedele`): drop rows with a run of
   consecutive missing values longer than `half` on both sides.
3. **Outlier removal**: `condextr` (point-by-point weighted
   conditional extremum) or `percoutl` (traditional percentile).
4. **Short-period interpolation** (`shorvalu`): fill remaining
   short gaps from nearby valid values.

Steps 1--3 are wrapped by `dataprep()` for one-call use.

## Why `obsedele` changed in 0.1.6

The 0.1.5 implementation collapsed all selected columns into one
long vector before computing missing runs, which merged runs across
columns and over-deleted boundary rows. The 0.1.6 C++ version
scans each column independently and keeps any row whose NA run is
shorter than `half` on at least one side. See `?obsedele` for the
formal definition.

The example below shows the new behaviour on a small synthetic
dataset.

```{r}
df <- data.frame(
  date  = as.POSIXct("2024-01-01 00:00:00", tz = "UTC") + 0:19 * 600,
  group = rep(1L, 20),
  x     = c(1, NA, NA, NA, 5, NA, NA, NA, NA, NA,
            1, NA, NA, NA, 5, NA, NA, NA, NA, NA),
  y     = c(NA, 1, NA, NA, 2, NA, NA, NA, NA, NA,
            NA, 1, NA, NA, 2, NA, NA, NA, NA, NA)
)
nrow(df)
obsedele(df, cols = c("x", "y"), group = "group", half = 2)
nrow(obsedele(df, cols = c("x", "y"), group = "group", half = 2))
```

## `condextr` vs `percoutl`

* `percoutl()` is a single-pass percentile threshold: every value
  above the top quantile or below the bottom quantile is set to NA.
  It is fast but has no notion of magnitude.
* `condextr()` adds two safeguards: a percentile **error margin**
  (`top.error`, `bottom.error`) and a **magnitude margin**
  (`top.magnitude`, `bottom.magnitude`). Only the extreme point of
  a window is removed if it exceeds the combined threshold. This
  preserves a wider dynamic range while still removing the most
  extreme values.

On `data1`:

```{r fig.width=6, fig.height=3.5}
par(mfrow = c(1, 2))
boxplot(data1$Nucleation, main = "raw",  ylab = "Nucleation")
boxplot(percoutl(data1, cols = "Nucleation")$Nucleation,
        main = "percoutl", ylab = "")
```

```{r fig.width=6, fig.height=3.5}
par(mfrow = c(1, 2))
boxplot(data1$Nucleation, main = "raw",  ylab = "Nucleation")
boxplot(condextr(data1, cols = "Nucleation",
                 interval = 5, times = 3)$Nucleation,
        main = "condextr", ylab = "")
```

## End-to-end example on the full `data` dataset

The `data1` subset used so far has only seven columns and very few
missing values, so it does not exercise `varidele`, `obsedele`, and
`condextr` in a realistic way. The full `data` table has 7,640 rows
and 65 columns (a time column, a grouping column, and 60 particle
size bins). We use it here to show how the pipeline changes the
shape of the data.

### A two-month slice

To keep the vignette fast we take two months (January 2020 and
July 2020) as a representative slice.

```{r}
data_slice <- data[data$monthyear %in% c("January 2020", "July 2020"), ]
num_cols   <- which(sapply(data_slice, is.numeric))
dim(data_slice)
length(num_cols)
```

#### Before cleaning

```{r fig.width=7, fig.height=4, warning=FALSE}
percplot(data_slice, cols = num_cols)
```

The curves above are the top and bottom percentiles of every size
bin, plotted on a log-log scale. Long, flat tails at the low and
high end of the size distribution indicate the presence of outliers
and long stretches of missing data.

#### Observation deletion

```{r}
step1 <- obsedele(data_slice, cols = num_cols, group = 4)
nrow(step1)
```

Rows whose selected size bins contain a gap longer than `half`
minutes on both sides are removed.

#### Conditional extremum outlier removal

```{r}
step2 <- condextr(step1, cols = num_cols, group = 4,
                  interval = 10, times = 10)
nrow(step2)
```

`condextr` combines a percentile threshold with an error margin and
a magnitude margin, and removes only the single most extreme value
in each window. Unlike `percoutl`, it does not flatten the tails.

#### After cleaning

```{r fig.width=7, fig.height=4, warning=FALSE}
num_cols_step2 <- which(sapply(step2, is.numeric))
percplot(step2, cols = num_cols_step2)
```

Compare with the earlier figure: the top and bottom percentile
curves are smoother, the extreme tails are shorter, and the
interquartile band is tighter. The number of observations in each
percentile label (`n =` in the legend) is smaller than before
because whole rows have been dropped.

#### Short-period interpolation

```{r}
step3 <- shorvalu(step2, cols = num_cols_step2)
sum(is.na(step2[, num_cols_step2])) - sum(is.na(step3[, num_cols_step2]))
```

`shorvalu` fills remaining short gaps (within `intervals = 30`
minutes) from the nearest valid values. The remaining `NA`s are in
gaps that are too long to interpolate.

### Full dataset

The same pipeline on the full 7,640-row `data` takes a few minutes
and is not run here:

```{r eval=FALSE}
cleaned <- dataprep(data,
                    cols       = which(sapply(data, is.numeric)),
                    group      = 4,
                    interval   = 10,
                    times      = 10,
                    intervals  = 30)
dim(cleaned)
```

Inside `dataprep()`, `varidele` drops columns whose missing
fraction exceeds `fraction`, `obsedele` and `condextr` run as
above, and `shorvalu` fills the remaining short gaps. The result
keeps the leading columns (`date`, `tconc`, `TPNC`, `monthyear`)
and the size bins that survived `varidele`.

## One-call pipeline: `dataprep()`

For quick exploration the four steps are wrapped in a single call:

```{r}
res <- dataprep(
  data1,
  cols     = c("Nucleation", "Aitken", "Accumulation"),
  group    = "monthyear",
  interval = 5,
  times    = 3,
  half     = 30
)
dim(res)
```

The arguments mirror the individual steps:

* `cols` selects the numeric variables to process.
* `group` selects the grouping column used by `obsedele` and
  `condextr`.
* `interval` and `times` control how aggressively `condextr` marks
  outliers between two observation-deletion rounds.
* `fraction` sets the missing-fraction cutoff for `varidele`.
* `half` and `by` define the consecutive-missing window for
  `obsedele`.
* `intervals` sets the maximum gap that `shorvalu` will fill.

## Inspecting the plan without running it

`dry_run()` reports what `varidele` / `obsedele` / `detect_outliers`
would remove without touching the data.

```{r}
report <- dry_run(
  data1,
  cols       = c("Nucleation", "Aitken", "Accumulation"),
  steps      = c("varidele", "obsedele", "outlier"),
  fraction   = 0.5
)
str(report, max.level = 2)
```

## Missing-value diagnostics

`na_diagnose()` summarises missing runs per variable:

```{r}
na_diagnose(data1, cols = 3:7)
```

## Additional cleaning helpers

| Function | Purpose |
|---|---|
| `detect_outliers()` | IQR / MAD / percentile masks |
| `winsorize()` | cap extreme values instead of removing them |
| `phys_filter()` | filter by physical bounds |
| `filter_high_cor()` | drop highly correlated variables |
| `filter_low_var()` | drop near-constant variables |
| `deduplicate()` | exact / fuzzy duplicate removal |
| `validate_data()` | rule-based validation |
| `balance_panel()` | balance an unbalanced panel |

```{r}
# Winsorize instead of removing
head(winsorize(data1, cols = "Nucleation")$Nucleation)

# Drop a constant column and a near-constant column
demo <- data.frame(
  id    = 1:100,
  const = rep(5, 100),
  noise = rnorm(100, sd = 0.005)
)
names(filter_low_var(demo, cutoff = 0.001))
```

## Session info

```{r}
sessionInfo()
```
