---
title: "dataprep: descriptive statistics and diagnostic plots"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: descriptive statistics and diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center"
)
```

```{r setup}
library(dataprep)
library(ggplot2)
library(scales)
```

## Overview

`dataprep` provides two families of plotting helpers:

* `descplot()` -- descriptive statistics (`n`, `na`, `mean`, `sd`,
  `median`, `trimmed`, `min`, `max`, `IQR`) computed in C++ and
  displayed as line or bar charts.
* `percplot()` -- top and bottom percentile curves, useful for
  detecting heavy tails and percentile-based outlier cutoffs.

Both share the same interface style: a data frame, a numeric range,
and optional grouping. Use `data1` (7,640 rows x 7 columns) for
quick demos, and `data` (7,640 x 65) for full-size examples.

## Descriptive statistics

### Line plot, numeric variable names

When variable names are essentially numeric (e.g. particle
diameters such as `3.16`, `3.55`, ...), `descplot()` draws a line
plot with a log-scaled x axis.

```{r fig.width=6, fig.height=4}
descplot(data1, cols = 3:7)
```

### Selected statistics

Pass a subset of statistics by index or by name to focus the plot.

```{r fig.width=6, fig.height=3}
descplot(data1, cols = 3:7,
         stats = c("na", "min", "max", "IQR"))
```

### Bar chart, character variable names

When variable names are character (e.g. aerosol mode names
`Nucleation`, `Aitken`, `Accumulation`), `descplot()` falls back
to a bar chart.

```{r fig.width=6, fig.height=5}
descplot(data1, cols = 3:7) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30,
                                        hjust = 1, vjust = 1.1))
```

### Control facet layout

```{r fig.width=6, fig.height=2.5}
descplot(data1, cols = 3:7, stats = c("min", "max", "IQR")) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30,
                                        hjust = 1, vjust = 1.1))
```

### Full-size data

```{r fig.width=7, fig.height=5, eval=FALSE}
descplot(data, cols = 5:65)
```

## Percentile curves

### Full percentile range

`percplot()` computes the extreme percentiles (0 to 0.5 and 99.5 to
100 by default) and draws them against the variable axis. This is
the visual companion to `condextr()` and `percoutl()`.

```{r fig.width=6, fig.height=4}
percplot(data1, cols = 3:7, group = 2)
```

### Top percentiles only

```{r fig.width=6, fig.height=2.5}
percplot(data1, cols = 3:7, group = 2, part = "top")
```

### Bottom percentiles only

```{r fig.width=6, fig.height=2.5}
percplot(data1, cols = 3:7, group = 2, part = "bottom")
```

### Numeric axis control

For numeric variable names, the x axis can be forced to linear
scale with `num_xaxis = "numeric"`.

```{r fig.width=6, fig.height=5}
percplot(data1, cols = 3:7, group = 2, num_xaxis = "numeric") +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30,
                                        hjust = 1, vjust = 1.1))
```

## Combining with `ggplot2`

Both `descplot()` and `percplot()` return `ggplot` objects, so all
usual `ggplot2` layers apply.

```{r fig.width=6, fig.height=3.5}
percplot(data1, cols = 3:7, group = 2) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "Percentile curves by month",
                x = "Variable", y = "Value")
```

## Session info

```{r}
sessionInfo()
```