---
title: "dataprep: cleaning pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: cleaning pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center"
)
```

```{r setup}
library(dataprep)
set.seed(1)
```

## Overview

The 0.1.6 cleaning pipeline consists of four sequential steps:

1. **Variable deletion** (`varidele`): drop columns whose missing
   fraction is above a threshold.
2. **Observation deletion** (`obsedele`): drop rows with a run of
   consecutive missing values longer than `half` on both sides.
3. **Outlier removal**: `condextr` (point-by-point weighted
   conditional extremum) or `percoutl` (traditional percentile).
4. **Short-period interpolation** (`shorvalu`): fill remaining
   short gaps from nearby valid values.

Steps 1--3 are wrapped by `dataprep()` for one-call use.

## Why `obsedele` changed in 0.1.6

The 0.1.5 implementation collapsed all selected columns into one
long vector before computing missing runs, which merged runs across
columns and over-deleted boundary rows. The 0.1.6 C++ version
scans each column independently and keeps any row whose NA run is
shorter than `half` on at least one side. See `?obsedele` for the
formal definition.

The example below shows the new behaviour on a small synthetic
dataset.

```{r}
df <- data.frame(
  date  = as.POSIXct("2024-01-01 00:00:00", tz = "UTC") + 0:19 * 600,
  group = rep(1L, 20),
  x     = c(1, NA, NA, NA, 5, NA, NA, NA, NA, NA,
            1, NA, NA, NA, 5, NA, NA, NA, NA, NA),
  y     = c(NA, 1, NA, NA, 2, NA, NA, NA, NA, NA,
            NA, 1, NA, NA, 2, NA, NA, NA, NA, NA)
)
nrow(df)
obsedele(df, cols = c("x", "y"), group = "group", half = 2)
nrow(obsedele(df, cols = c("x", "y"), group = "group", half = 2))
```

## `condextr` vs `percoutl`

* `percoutl()` is a single-pass percentile threshold: every value
  above the top quantile or below the bottom quantile is set to NA.
  It is fast but has no notion of magnitude.
* `condextr()` adds two safeguards: a percentile **error margin**
  (`top.error`, `bottom.error`) and a **magnitude margin**
  (`top.magnitude`, `bottom.magnitude`). Only the extreme point of
  a window is removed if it exceeds the combined threshold. This
  preserves a wider dynamic range while still removing the most
  extreme values.

On `data1`:

```{r fig.width=6, fig.height=3.5}
par(mfrow = c(1, 2))
boxplot(data1$Nucleation, main = "raw",  ylab = "Nucleation")
boxplot(percoutl(data1, cols = "Nucleation")$Nucleation,
        main = "percoutl", ylab = "")
```

```{r fig.width=6, fig.height=3.5}
par(mfrow = c(1, 2))
boxplot(data1$Nucleation, main = "raw",  ylab = "Nucleation")
boxplot(condextr(data1, cols = "Nucleation",
                 interval = 5, times = 3)$Nucleation,
        main = "condextr", ylab = "")
```

## One-call pipeline: `dataprep()`

```{r}
res <- dataprep(
  data1,
  cols     = c("Nucleation", "Aitken", "Accumulation"),
  group    = "monthyear",
  interval = 5,
  times    = 3,
  half     = 30
)
dim(res)
```

## Inspecting the plan without running it

`dry_run()` reports what `varidele` / `obsedele` / `detect_outliers`
would remove without touching the data.

```{r}
report <- dry_run(
  data1,
  cols       = c("Nucleation", "Aitken", "Accumulation"),
  steps      = c("varidele", "obsedele", "outlier"),
  fraction   = 0.5
)
str(report, max.level = 2)
```

## Missing-value diagnostics

`na_diagnose()` summarises missing runs per variable:

```{r}
na_diagnose(data1, cols = 3:7)
```

## Additional cleaning helpers

| Function | Purpose |
|---|---|
| `detect_outliers()` | IQR / MAD / percentile masks |
| `winsorize()` | cap extreme values instead of removing them |
| `phys_filter()` | filter by physical bounds |
| `filter_high_cor()` | drop highly correlated variables |
| `filter_low_var()` | drop near-constant variables |
| `deduplicate()` | exact / fuzzy duplicate removal |
| `validate_data()` | rule-based validation |
| `balance_panel()` | balance an unbalanced panel |

```{r}
# Winsorize instead of removing
head(winsorize(data1, cols = "Nucleation")$Nucleation)

# Drop a constant column and a near-constant column
demo <- data.frame(
  id    = 1:100,
  const = rep(5, 100),
  noise = rnorm(100, sd = 0.005)
)
names(filter_low_var(demo, cutoff = 0.001))
```

## Session info

```{r}
sessionInfo()
```
