---
title: "dataprep: upgrading from 0.1.5 to 0.1.6"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: upgrading from 0.1.5 to 0.1.6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(dataprep)
```

## What changed

Three behaviour changes affect row and column counts. All three are
bug fixes, but each one changes the output on real data. If your
downstream analysis depends on exact row counts, read the
quantified comparison in the last section before upgrading.

1. **`obsedele()` now scans each column independently.** The 0.1.5
   implementation collapsed all selected columns into one long
   vector before computing missing runs; this merged NA runs
   across columns and over-deleted boundary rows.

2. **The `half`-minute boundary is now inclusive.** Rows whose
   nearest anchor is exactly `half` minutes away are retained
   (`within half minutes` is a `<=` condition).

3. **`optisolu()` no longer crashes with `cores > 16`.** The 0.1.5
   `parallel::makeCluster()` path exhausted memory when the
   worker processes each received a full copy of the input. The
   0.1.6 implementation shares read-only data across workers, so
   `cores = 64` and `cores = NULL` (automatic) are both safe.

The full description is in `news(package = "dataprep")`. The design
reasoning behind the pipeline is in
`vignette("dataprep-philosophy")`.

## Minimal reproduction of the boundary change

Consider five observations sampled every 10 minutes, with a valid
value only at the two ends:

```{r}
df <- data.frame(
  date = as.POSIXct("2024-01-01 00:00:00", tz = "UTC") + 0:4 * 600,
  x    = c(1, NA, NA, NA, 5)
)

obsedele(df, cols = "x", half = 30)
```

The three interior rows are each 10, 20, or 30 minutes from the
nearest valid anchor. Under 0.1.5 the third row was deleted because
the comparison was strict (`dl > half`); under 0.1.6 it is retained
because `dl == half` satisfies `within half minutes`.

## Minimal reproduction of the multi-column change

Consider two channels with mutually exclusive missing runs:

```{r}
df <- data.frame(
  date = as.POSIXct("2024-01-01 00:00:00", tz = "UTC") + 0:9 * 600,
  x    = c(1, NA, NA, NA, NA, NA, NA, NA, NA, 5),
  y    = c(NA, NA, NA, NA, 2, NA, NA, NA, NA, NA)
)
nrow(obsedele(df, cols = c("x", "y"), half = 60))
```

Under 0.1.5, the x and y NA runs were merged before computing the
run length, so the interior rows survived even when only one
channel was locally anchored. Under 0.1.6 the two channels are
checked independently: a row is deleted when **every** selected
column has a run longer than `half` minutes on **both** sides.

## Implementation evolution

The retention criterion has been stable across releases, but the
implementation has improved steadily. The table below summarises
the three generations:

| Release | Strategy | Complexity |
|---|---|---|
| 0.1.0 | Borrowed running mean: expand the series onto a regular grid with `tidyr::complete()`, compute a 59-minute centred moving average on a temporary column, and use its emptiness pattern to flag long runs. | O(grid length) per subset |
| 0.1.5 | Run-length encoding: use `data.table::rleid()` and `rowid()` to collapse consecutive NAs into runs, then compare each run length to the number of grid points covered by `half` minutes. | O(n) time, O(n) temporary storage |
| 0.1.6 | Anchor scan: for each missing value, look up the nearest non-missing anchor on each side and compare the two time distances directly. No grid, no run-length state. | O(n) time, O(1) extra allocation per column |

Each generation produces the same deletion decision on the same
input, but the constant factors shrink. The 0.1.6 anchor scan is
the first version that is fast enough to run interactively on
full-year data: on SMEAR I Varrio 2025 (49,422 rows × 61 numeric
channels), `obsedele()` now runs in 20 ms against 11.2 s in 0.1.5
(a 560× speed-up). See `vignette("dataprep-performance")` for the
full benchmark.

## Quantified effect on a full-year dataset

On SMEAR I Varrio 2025 (49,422 rows × 61 numeric channels,
10-minute sampling), running the same pipeline with the same
parameters:

| Stage | 0.1.5 | 0.1.6 | Δ |
|---|---|---|---|
| `varidele` | 25 columns deleted | 25 columns deleted | 0 |
| `obsedele` | 1,494 rows deleted | 1,496 rows deleted | +2 |
| `condextr` | 1,868 rows deleted | 1,863 rows deleted | −5 |
| `shorvalu` | 50,376 NAs filled | 50,387 NAs filled | +11 |
| `dataprep` final | 46,060 rows | 46,063 rows | **+3** |

Net change: 0.006% of the input. The five rows that differ
between versions all sit at run boundaries where the anchor
distance is within one sampling interval of `half` minutes.

## Where the difference comes from

The five differing rows fall into two groups:

**Rows kept by 0.1.6, deleted by 0.1.5 (3 rows).** These are rows
whose nearest anchor is exactly `half` minutes away. Under the
0.1.5 strict comparison (`dl > half`) they were deleted; under the
0.1.6 inclusive comparison (`dl >= half`) they are retained. Each
of these rows has a valid anchor within one sampling interval of
`half`, so retaining them is consistent with the physical
constraint described in `vignette("dataprep-philosophy")`.

**Rows deleted by 0.1.6, kept by 0.1.5 (3 rows, overlapping with
the above).** These are rows that pass the check on some columns
but fail on at least one. Under 0.1.5 the merge-columns approach
effectively widened the anchor window on these rows; under 0.1.6
each column is checked independently, so the row is deleted. These
rows would have been interpolated across a gap longer than `half`
minutes in at least one channel, which contradicts the design.

The net result is 3 additional rows in the final output.

## Migration checklist

- [ ] Re-run any downstream analysis that uses the output of
      `obsedele()`, `condextr()`, `percoutl()`, or `dataprep()`.
      The three-row difference on SMEAR I Varrio 2025 is
      representative of the magnitude you should expect on other
      datasets (well under 0.1% of rows).
- [ ] If you use `prep_fit()` / `prep_transform()`, re-fit the
      plan on the new training data. The stored thresholds are not
      affected by the change, but the row indices that fed into
      them are.
- [ ] If you were limiting `optisolu()` to `cores <= 16` as a
      workaround for the 0.1.5 crash, remove the cap. 0.1.6
      accepts up to 64.
- [ ] If downstream behaviour depends on specific boundary rows,
      verify the difference with `dplyr::anti_join()` between the
      0.1.5 and 0.1.6 outputs. The example below shows how.

```{r eval=FALSE}
result_015 <- dataprep::dataprep(data, cols = 5:65, group = 4)
result_016 <- dataprep(data, cols = 5:65, group = 4)

kept_only_by_016 <- dplyr::anti_join(result_016, result_015, by = "date")
kept_only_by_015 <- dplyr::anti_join(result_015, result_016, by = "date")
```

## What did not change

For users who only call `melt()` and `dcast()`, or who only use
`dataprep` for descriptive statistics (`descdata`, `na_diagnose`,
`percdata`, `percplot`, `descplot`), there is no behaviour change.
Those functions have been re-implemented in 0.1.6 for speed
(`melt` and `dcast`) or reorganised internally (`data_report`,
`dry_run`), but the output on every tested input is byte-identical
to 0.1.5 except where noted in the `news` file.

## Where to go next

* **Design philosophy** — `vignette("dataprep-philosophy")`.
  Why the pipeline has the shape it does.
* **Cleaning pipeline walkthrough** — `vignette("dataprep-cleaning")`.
  Step-by-step execution on a real dataset.
* **Performance and cross-engine consistency** —
  `vignette("dataprep-performance")`. Full benchmark tables and
  8-engine consistency checks.
* **Fast reshaping** — `vignette("dataprep-melt-dcast")`.

## Session info

```{r}
sessionInfo()
```