---
title: "dataprep: a leakage-free preprocessing workflow"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: a leakage-free preprocessing workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center"
)
```

```{r setup}
library(dataprep)
set.seed(1)
```

## The data-leakage problem

Standard preprocessing steps such as outlier detection, imputation,
and scaling are often implemented as one-shot functions. If you
apply them to training and test data **separately**, the test set
ends up using its own statistics, which leaks information from the
test set into the pipeline.

`dataprep` 0.1.6 solves this with a two-step interface:

* `prep_fit()` estimates every parameter (missing fraction, outlier
  thresholds, imputation method, scaling centre/scale) from the
  **training data only**, and returns a `prep_plan`.
* `prep_transform()` applies the plan to **new data** without
  re-estimating anything.

## A minimal example

```{r}
train <- data1[1:5000, c("date", "monthyear",
                         "Nucleation", "Aitken", "Accumulation")]
test  <- data1[5001:6000, c("date", "monthyear",
                            "Nucleation", "Aitken", "Accumulation")]

plan <- prep_fit(
  train,
  cols       = 3:5,
  group      = 2,
  steps      = c("varidele", "outlier", "impute", "scale"),
  fraction   = 0.5,
  method_outlier = "iqr",
  method_impute  = "linear",
  scale_method   = "zscore"
)

str(plan, max.level = 2)
```

Apply the plan to the test set:

```{r}
test_clean <- prep_transform(plan, test)
head(test_clean)
```

## What is stored in the plan

```{r}
names(plan)
names(plan$params)
```

* `params$varidele_keep` -- logical mask of columns to keep
* `params$outlier_thresholds` -- lower / upper bounds per column
  (or per column-group)
* `params$impute_method` -- imputation method string
* `params$scale_center`, `params$scale_scale` -- per-column
  centring and scaling constants
* `data_info` -- original column names and indices, used to
  realign columns when `test` is fed in with a different order

## Adding a custom step

`prep_fit()` accepts an ordered `steps` vector. Any subset of the
following is allowed:

```{r eval=FALSE}
steps = c("varidele", "obsedele", "outlier",
          "impute", "scale")
```

* `varidele` -- drop columns whose training-set missing fraction
  is above `fraction`.
* `obsedele` -- drop rows with long consecutive NA runs.
* `outlier` -- detect and replace outliers with NA using IQR, MAD,
  or percentile thresholds estimated on the training set.
* `impute` -- fill remaining NAs using LOCF, NOCB, linear, mean,
  or median.
* `scale` -- z-score, min-max, robust, centre, or scale.

## Workflow with `dataprep()`

For exploratory analysis where leakage is not a concern, the
one-call `dataprep()` wrapper chains the four standard steps:

```{r}
res <- dataprep(
  data1,
  cols     = c("Nucleation", "Aitken", "Accumulation"),
  group    = "monthyear",
  interval = 5,
  times    = 3
)
dim(res)
```

## Reporting

`data_report()` prints a compact quality summary:

```{r}
data_report(data1, cols = 3:7, verbose = TRUE)
invisible(data_report(data1, cols = 3:7))
```

## Full applied workflow

```{r eval=FALSE}
# 1. Inspect
data_report(data1, cols = 3:7, verbose = TRUE)

# 2. Fit a plan on the training split
plan <- prep_fit(
  train,
  cols     = 3:5,
  group    = 2,
  fraction = 0.5
)

# 3. Apply the same plan to the test split
test_clean <- prep_transform(plan, test)

# 4. Model on the cleaned training set,
#    predict on the cleaned test set
fit  <- lm(Nucleation ~ Aitken + Accumulation,
           data = plan$final_data)
pred <- predict(fit, newdata = test_clean)
```

## Session info

```{r}
sessionInfo()
```