---
title: "dataprep: descriptive statistics and diagnostic plots"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: descriptive statistics and diagnostic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "95%"
)
```

```{r}
library(dataprep)
library(ggplot2)
```

## Overview

`dataprep` provides two plotting helpers, both built on top of the
C++ descriptive backends:

* **`descplot()`** — descriptive statistics (`n`, `na`, `mean`, `sd`,
  `median`, `trimmed`, `min`, `max`, `IQR`) computed in C++ and
  displayed as a line or bar chart.

* **`percplot()`** — top and bottom percentiles, useful for detecting
  heavy tails and for comparing with percentile-based outlier cutoffs.

Both functions share a common calling style: a data frame, a numeric
column range, and optional faceting. They differ in their extra
arguments — `descplot()` takes `stats` and `first`; `percplot()`
takes `group`, `diff`, and `part`.

Both also choose the geometry from the **column names** of the
selected variables:

* numeric names (e.g. particle diameters `3.16`, `3.55`, ...)
  → line chart, optionally on a log-scaled x axis;
* character names (e.g. `Nucleation`, `Aitken`, `Accumulation`)
  → bar chart.

The examples below use two bundled data sets:

* `data` — 7,640 rows × 65 columns; columns 5–65 are named by
  particle diameter (numeric names). Missingness is uneven: a few
  channels (`1` to `2.51`, `794`, `891`, `1000`) are entirely `NA`,
  while several others (e.g. `2.82`, `3.16`) are missing for whole
  months.
* `data1` — 7,640 rows × 7 columns; columns 3–7 are `Nucleation`,
  `Aitken`, `Accumulation`, `tconc`, `TPNC` (character names).

## Descriptive statistics with `descplot()`

### Numeric variable names: line chart

When the selected columns have numeric names, `descplot()` draws a
line chart with a log-scaled x axis by default.

```{r}
descplot(data, cols = 5:65)
```

### Character variable names: bar chart

When the selected columns have character names, `descplot()` falls
back to a bar chart.

```{r}
descplot(data1, cols = 3:7) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30,
                                        hjust = 1, vjust = 1.1))
```

### Selecting statistics

Pass a subset of statistics by index or by name.

```{r}
descplot(data, cols = 5:65,
         stats = c("na", "min", "max", "IQR"))
```

`stats` accepts both numeric indices (`1:9`) and character names
(`"na"`, `"min"`, `"max"`, `"IQR"`), and can mix the two.

### Controlling the facet layout

Use `ncol` to control how many facets are placed per row.

```{r}
descplot(data1, cols = 3:7, stats = c("min", "max", "IQR"),
         ncol = 2) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30,
                                        hjust = 1, vjust = 1.1))
```

### The underlying table

If only the numbers are needed, call `descdata()` directly. It
returns one row per variable and one column per statistic.

```{r}
descdata(data1, cols = 3:7, stats = c(2, 3, 4, 7:9))
```

## Percentiles with `percplot()`

### Numeric variable names: line chart

With numeric column names and `part = "both"` (the default),
`percplot()` draws a line chart. Because bottom and top percentiles
span a wide dynamic range, the y axis is automatically placed on a
log10 scale.

```{r}
percplot(data, cols = 5:65, group = 4)
```

### Missing values in the long table

Internally, `percplot()` reshapes the wide output of `percdata()`
into a long table whose rows are `(group, percentile, variable)`
triples. If a variable has no non-missing observation inside a
given group, `quantile(na.rm = TRUE)` returns `NA` for every
requested probability, so all the rows for that `(group, variable)`
pair have `value = NA`. `ggplot2` drops those rows and prints a
warning similar to

```
Removed 288 rows containing missing values (`geom_line()`).
```

Here `288` is the count of such rows in the **long** table, not a
count of dropped columns. Because the default `part = "both"`
returns 12 percentiles, 288 rows correspond to `288 / 12 = 24`
`(group, variable)` pairs that are entirely missing within their
group. Two mechanisms produce them:

1. a few channels (`1`–`2.51`, `794`–`1000`) are entirely `NA`
   throughout the record, and
2. several channels (`2.82`, `3.16`, `3.55`, …) are missing for
   whole months, so within that month they have no value at all.

The visible lines are still correct. To count the `NA` cells
directly:

```{r}
pct <- percdata(data, cols = 5:65, group = 4)
sum(is.na(pct[, -(1:2), drop = FALSE]))
```

Restricting to a contiguous range of channels that are present in
every month removes the warning:

```{r}
percplot(data, cols = 27:61, group = 4)
```

### Top or bottom percentiles only

Use `part = "top"` or `part = "bottom"` to keep only one tail.

```{r}
percplot(data, cols = 27:61, group = 4, part = "top")
```

```{r}
percplot(data, cols = 27:61, group = 4, part = "bottom")
```

### Character variable names: bar chart

When the column names are character, `percplot()` draws grouped bar
charts instead of line charts.

```{r}
percplot(data1, cols = 3:7, group = 2) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30,
                                        hjust = 1, vjust = 1.1))
```

### Numeric axis control

For numeric column names, `num_xaxis` controls the x axis:

* `"auto"` (default for `percplot()`) — log scale if the column
  names are evenly spaced on a log scale with a range ratio of at
  least 1000, otherwise linear scale.
* `"log"` / `"numeric"` — force the corresponding scale.
* `"character"` / `"factor"` / `"keep"` / `FALSE` — keep the column
  names as factor levels and draw a bar chart even if the names look
  numeric.

Note that `descplot()` defaults to `num_xaxis = "log"`, while
`percplot()` defaults to `"auto"`.

```{r}
percplot(data, cols = 27:61, group = 4, num_xaxis = "numeric")
```

### The underlying table

`percdata()` returns the same table that `percplot()` draws.

```{r}
percdata(data1, cols = 3:7, group = 2, part = "top")
```

## A diagnostic workflow

A typical diagnostic workflow combines `data_report()`,
`na_diagnose()`, and the two plot families:

```{r, eval = FALSE}
# 1. Overview of the whole table
data_report(data, cols = 5:65)

# 2. Per-column NA run statistics
na_diagnose(data, cols = 5:65)

# 3. Descriptive statistics of the raw data
descplot(data, cols = 5:65)

# 4. Percentile plots of the raw data
percplot(data, cols = 5:65, group = 4)
```

After running `dataprep()` you can compare the raw and cleaned
versions in the same plot by stacking them with a `g` column:

```{r, eval = FALSE}
cleaned <- dataprep(data, cols = 5:65, group = 4)

percplot(
  rbind(
    transform(data[names(cleaned)], g = "original"),
    transform(cleaned,               g = "preprocessed")
  ),
  cols  = 5:ncol(cleaned),
  group = ncol(cleaned) + 1
)
```

## Where to go next

* **Design philosophy** — `vignette("dataprep-philosophy")`.
* **Cleaning pipeline walkthrough** —
  `vignette("dataprep-cleaning")`.
* **Performance and cross-engine consistency** —
  `vignette("dataprep-performance")`.
* **Upgrading from 0.1.5 to 0.1.7** —
  `vignette("dataprep-migration")`.
* **Fast reshaping** — `vignette("dataprep-melt-dcast")`.

## Session info

```{r}
sessionInfo()
```